# Introduction to R 

This is a brief introduction to R, focussing on basic programming concepts and best practices.

## Workflow

You can interact directly with the console, but it is considered best practice to write an .R file and run commands in the console from that script. This allows saving the code and running all at once with the `source` function. Remember to click Save often, so that you actually save your script!


Once you open an R script (or any file), an editor panel will appear in the top left. From there you can run the current line (or a selected block of code) by clicking on the `Run` button in the editor panel, or in the `Code` menu select `Run Lines`. There is also a `Re-run` button next to the `Run` button, which allows to run again the previous code block including any new modifications you have just made.

## Working directory

The working directory is a folder on your computer where R will look for data, save your plots, etc. To make your workflow easier, it is good practice to save everything related to one project in the same place, as it will save you time typing up computer paths or finding the files across your computer. For example, you could save your script and all the data for this exercises in a folder called "Intro-to-R". It is good practice to avoid spaces in file names because it can confuse R.

To find out where your working directory is now, run the function `getwd()`. If you want to change it, you can use `setwd("dir")`, where "dir" is the path to the folder that you want to use. Notice that the path must be written inside quotation marks. You'll see that in the console R runs the command for changing the working directory. 

Alternatively, you can set the working directory from the Files tab in the Files/Plot/Packages/Help pane. For this, navigate to the folder whence you want to read and save files or create a new one by clicking `New Folder`. Once you have located the folder, click on it to access it. Then, click `More` and select `Set As Working Directory`. 

**TRY IT!** Set the working directory to a folder in your Desktop called "Intro-to-R". 

## Variables and assignment

The assignment operator is `<-`. It stores whatever you write on the right to variables on
the left. Variable names can be made up of letters, numbers, underscores and periods. But, they cannot start with a number nor contain spaces. For instance:

```{r}
x <- 10
```

Notice that assigning a variable does not print its value, 
but if you look at the `Environment` panel, you will see the new variable there.
You can also check that it is stored by calling the variable:

```{r}
x 
```

**TRY IT!** Define a varible "y" with the value of 3. 

### Shortcuts

In RStudio, the assignment operator is written in a keystroke by hitting `Alt`+`-`.

A practical way of running code from a script by using a keyboard shortcut. In GNU/Linux and Windows, you can hit `Ctrl`+`Return`, on macOS the equivalent is `⌘ Command`+`Return`. This shortcut also allows running a selected block of code. 

You can alternate between the script and console panels with the `Ctrl`+`1` and `Ctrl`+`2` shortcuts.

From the console, you can go back to the previous command by hitting the `up` arrow. If you press the `up` arrow multiple times you will walk back through the history. You can navigate the history by pressing the `down` arrow as well. You can edit any command from there by using the `left` arrow or `delete` keys.

## Using variables

Variables can be used in calculations

```{r}
x + x
```

Variables can also be reassigned

```{r}
x <- 314
x
```

## Comments

In R, the `#` sign is a special character that indicates a comment. Anything that is written directly to the right of a `#` sign will be ignored by R. Comment your scripts often!

## Precedence

The order of operations follows the standard precedence (from highest to lowest):

* Parentheses `(`, `)`
* Exponents or roots: `^`
* Multiplication or division: `*`, `/`
* Addition or subtraction: `+`, `-`

## Mathematical functions

R comes with many built in mathematical functions. Functions are called by typing its name followed by open and closing parentheses. Whatever we write inside the parentheses are the function's arguments:

```{r}
sqrt(x)
```

## Vectorization

So far we assigned our variables with *scalars* (a 1 dimensional object with a length of 1). But, R is a vectorized language, which means that variables and functions can have vectors as values. A *vector* is a set of values, in a specific order, and of the same data type. 

Vectors are *atomic*, which means that they can only contain one class of data.
```{r}
rgb <- c("red", "green", "blue")
rgb
y <- 3:7
y
```

# check the type of data
```{r}
class(x)
class(rgb)
```

# Vectors can also be used as arguments to functions

```{r}
length(y) # count the number of elements in y
```

# But different classes cannot be mixed in the same object

```{r}
semiprimes <- c(4, 6, 9, "ten")
class(semiprimes)
```

The core idea behind vectorization is that operations can be applied at once 
to an entire set of values

```{r}
y * 3
```

**TRY IT!** Create a vector with a sequence of numbers using the function `seq`. Hint: try `?seq`.

## R Packages

Functions can be added to R with packages, which are stored on [CRAN](https://cran.r-project.org/web/packages/index.html) (the Comprehensive R Archive Network). 

To install a package you use the function `install.packages("packagename")`, where `packagename` is the name of the package you want to install. To see what packages are installed by type `installed.packages()`.

Packages are loaded with the `library("packagename")` function. 
